Detecting distributional differences between temporal granularities for exploratory time series analysis

### 2021-10-28 paper/main.Rmd
Working on sec 3

Pairwise distances could be within-facets or between-facets for $m \geq 2$. Figure \ref{fig:distance-explain} illustrates how they are defined. Pairwise distances are within-facets when $b_{k} = b_{k'}$, that is, between pairs of the form $(a_{j}b_{k}, a_{j'}b_{k})$ as shown in panel (3) of Figure \ref{fig:distance-explain}. If categories are ordered (like all temporal cyclic granularities), then only distances between pairs where $a_{j'} = (a_{j+1})$ are considered (panel (4)). Pairwise distances are between-facets when they are considered between pairs of the form $(a_{j}b_{k}, a_{j}b_{k'})$. There are a total of ${K \choose 2}J$ between-facet distances, and ${J \choose 2}K$ within-facet distances if they are ordered and $K(J-1)$ within-facet distances if they are unordered.

@@

### Adjusting for the number of comparisons {-}

@@

### _Permutation approach_ {-}

@@

### _Modeling approach_ {-}

@@

### _Combination approach_ {-}

@@

A cyclic granularity is referred to as "significant" if there is a significant distributional difference of the measured variable between different categories of the harmony. In this section, a selection criterion to choose significant harmonies is provided, thereby eliminating all harmonies that exhibit complete randomness in the measured variable. The distance measure $\wpd$ is used as a test statistic to test the null hypothesis that no harmony/cyclic granularity is significant. We select only those harmonies/cyclic granularities for which the test fails. They are then ranked basis how well they capture variation in the measured variable.

@@

A threshold and consequently a selection criterion is chosen using the notion of Randomization tests [@edgington2007randomization]. The data is permuted several times and $\wpd$ is computed for each of the permuted data sets to obtain the sampling distribution of $\wpd$ under the null hypothesis. If the null hypothesis is true, then $\wpd$ obtained from the original data set would be a likely value in the sampling distribution. But in case the null hypothesis is not true, then it is less probable that $\wpd$ obtained for the original data will be from the same distribution. This idea is utilized to come up with a threshold for selection, denoted by $\wpdsub{threshold}$, defined as the $99^{th}$ percentile of the sampling distribution. A harmony is selected if the value of $\wpd$ for that harmony is greater than the chosen threshold. The detailed algorithm for choosing a threshold and selection procedure (for two cyclic granularities) is listed as follows:

@@

- **Input:** All harmonies of the form $\{(A, B), A = \{ a_j: j = 1, 2, \dots, J\}, B = \{ b_k: k = 1, 2, \dots, K\}\}$, $\forall (A, B) \in H_{N_C}$.

@@

1. Fix harmony pair $(A, B)$.

2. Given the measured variable; $\{v_t: t=0, 1, 2, \dots, T-1\}$, $\wpd$ is computed and is represented by $\wpd^{A, B}_{obs}$.

3. From the original sequence a random permutation is obtained: $\{v_t^{1}: t=0, 1, 2, \dots, T-1\}$.

4. $\wpd$ is computed for the permuted sequence of the data and is represented by $\wpd^{A, B}_{1}$.

5. Steps (3) and (4) are repeated a large number of times $M$ (e.g. $M = 200$).

6. For each permutation, one $\wpd^{A, B}_{i}$ is obtained. Define $\wpdsub{sample} = \{\wpd^{A, B}_{1} \wpd^{A, B}_{2}, \dots, \wpd^{A, B}_{M}\}$.

7. Repeat Steps (1-6) for all harmony pairs $(A, B) \in H_{N_C}$ and stored $\wpdsub{sample}^{\text{all}}$.

8. $99^{th}$ percentiles of $\wpdsub{sample}^{\text{all}}$ is computed and stored in $\wpdsub{threshold99}$

7. If $\wpd^{A, B}_{obs} > \wpdsub{threshold99}$, harmony pair $(A, B)$ is selected, otherwise rejected.

Similarly, a harmony pair $(A, B)$ could be selected if $\wpd^{A, B}_{obs} > \wpdsub{threshold95}$ and $\wpd^{A, B}_{obs} > \wpdsub{threshold90}$, where $\wpdsub{threshold95}$ and $\wpdsub{threshold90}$ denote the $95^{th}$ and $90^{th}$ percentile of $\wpdsub{sample}^{\text{all}}$ respectively. A harmony selected using $99^{th}$, $95^{th}$ and $90^{th}$ threshold are tagged as \*\*\*, \*\*, \* respectively.

@@

The distribution of $\wpd$ is expected to be similar for all harmonies under the null hypothesis, since they have been adjusted for different number of categories for the harmonies or underlying distribution of the measured variable. Hence, the values of $\wpd$ for different harmonies are comparable and can be used to rank the significant harmonies. A higher value of $\wpd$ for a harmony indicates that higher maximum variation in the measured variable is captured through that harmony. Figure \ref{fig:null4by2} presents the results of $\wpd$ from the illustrative designs in Section \ref{sec:computation-wpd}. The value of $\wpd$ under null design (a) is the least, followed by (b), (c) and (d). This aligns with the principle of $\wpd$, which is expected to have lowest value for null designs and highest for designs of the form $\Dfx$ (d). Moreover, note the relative differences in $\wpd$ values between (b) and (c). The value of the tuning parameter $\lambda$ is set to $0.67$, which has resulted in giving more emphasis to differences in x-axis categories. Again consider \ref{fig:onegran-new}(a) and \ref{fig:onegran-new}(b) with a $\wpd$ value of 20.5 and 145 respectively. This is because there is more gradual increase across hours of the day than months of the year. If order of the categories is not considered, they result in a $\wpd$ value of $97.8$ and $161$ respectively, which follows from the fact that if we consider difference between any hours of the day, the magnitude will be much larger than if we consider difference between consecutive categories. Similarly, \ref{fig:onegran-new}(a) and (b) have $\wpd$ values of  $110.79$ and $125.82$ respectively. The ranking implies that the distributional differences are more prominent for the second household, as is also seen from the bigger fluctuations in the $90^{th}$ percentile than the first household.

@@

```{r plot-all-new3, fig.cap = "(ref:plot-all-new3", dependson= "null4by2"}

@@

```{r raw, fig.cap = "(ref:raw"}
```

@@

(ref:dist-new-same-scale-link) The distribution of $\wpdsub{perm}$ and $\wpdsub{glm-scaled}$ are overlaid to compare the location and scale across different $\nx$ and $\nf$. $\wpdnorm$ takes the value of $\wpdsub{perm}$ for lower levels, and $\wpdsub{glm-scaled}$ for higher levels to to alleviate the problem of computational time in permutation approaches. This is possible as the distribution of the adjusted measure looks similar for both approaches for higher levels.

```{r dist-new-same-scale-link, fig.cap = "(ref:dist-new-same-scale-link"}
```

>>>>

Pairwise distances could be within-facets or between-facets for $m = 2$. Figure \ref{fig:distance-explain} illustrates how they are defined. Pairwise distances are within-facets when $b_{k} = b_{k'}$, that is, between pairs of the form $(a_{j}b_{k}, a_{j'}b_{k})$ as shown in panel (3) of Figure \ref{fig:distance-explain}. If categories are ordered (like all temporal cyclic granularities), then only distances between pairs where $a_{j'} = (a_{j+1})$ are considered (panel (4)). Pairwise distances are between-facets when they are considered between pairs of the form $(a_{j}b_{k}, a_{j}b_{k'})$. There are a total of ${K \choose 2}J$ between-facet distances, and ${J \choose 2}K$ within-facet distances if they are ordered and $K(J-1)$ within-facet distances if they are unordered.

@@

## Adjusting for the number of comparisons

@@

### Permutation approach {-}

@@

### Modeling approach {-}

@@

### Combination approach {-}

@@

A cyclic granularity is referred to as "significant" if there is a significant distributional difference of the measured variable between different categories of the harmony. In this section, a selection criterion to choose significant harmonies is provided, thereby eliminating all harmonies that exhibit non-significant differences in the measured variable. The distance measure $\wpd$ is used as a test statistic to test the null hypothesis that no harmony/cyclic granularity is significant. We select only those harmonies/cyclic granularities for which the test fails. They are then ranked based on how well they capture variation in the measured variable.

@@

A threshold (and consequently a selection criterion) is chosen using the notion of randomization tests [@edgington2007randomization]. The data is permuted several times and $\wpd$ is computed for each of the permuted data sets to obtain the sampling distribution of $\wpd$ under the null hypothesis. If the null hypothesis is true, then $\wpd$ obtained from the original data set would be a likely value in the sampling distribution. But in case the null hypothesis is not true, then it is less probable that $\wpd$ obtained for the original data will be from the same distribution. This idea is utilized to come up with a threshold for selection, denoted by $\wpdsub{threshold}$, defined as the $99^{th}$ percentile of the sampling distribution. A harmony is selected if the value of $\wpd$ for that harmony is greater than the chosen threshold. The detailed algorithm for choosing a threshold and selection procedure (for two cyclic granularities) is listed as follows:

@@

- **Input:** All harmonies of the form $\{(A, B),~ A = \{ a_j: j = 1, 2, \dots, J\},~ B = \{ b_k: k = 1, 2, \dots, K\}\}$, $\forall (A, B) \in H_{N_C}$.

@@

1. For each harmony pair $(A, B) \in H_{N_C}$, the following steps are taken.

    a. Given the measured variable; $\{v_t: t=0, 1, 2, \dots, T-1\}$, $\wpd$ is computed and is represented by $\wpd^{A, B}_{obs}$.
    b. For $i=1,\dots,m$, randomly permute the original time series: $\{v_t^{i}: t=0, 1, 2, \dots, T-1\}$ and compute $\wpd^{A, B}_{i}$ from $\{v_t^{i}\}$.
    c. Define $\wpdsub{sample} = \{\wpd^{A, B}_{1}, \dots, \wpd^{A, B}_{M}\}$.

2. Stack the $\wpdsub{sample}$ vectors as $\wpdsub{sample}^{\text{all}}$ and compute its $p = 100(1-\alpha)$ percentiles as $\wpdsub{threshold\emph{p}}$.
3. If $\wpd^{A, B}_{obs} > \wpdsub{threshold\emph{p}}$, harmony pair $(A, B)$ is selected at the $1-p/100$ level, otherwise rejected.
4. Harmonies selected using the $99^{th}$, $95^{th}$ and $90^{th}$ thresholds are tagged as \*\*\*, \*\*, \* respectively.

@@

The distribution of $\wpd$ is expected to be similar for all harmonies under the null hypothesis, since they have been adjusted for different number of categories for the harmonies or underlying distribution of the measured variable. Hence, the values of $\wpd$ for different harmonies are comparable and can be used to rank the significant harmonies. A higher value of $\wpd$ for a harmony indicates that higher maximum variation in the measured variable is captured through that harmony.

```{r plot-all-new3, fig.cap = "(ref:plot-all-new3)", dependson= "null4by2"}

\todo[inline]{Plot missing and plot-all-new3 code is not relevant.}

Figure \ref{fig:plot-all-new3} presents the results of $\wpd$ from the illustrative designs in Section \ref{sec:computation-wpd}. The value of $\wpd$ under null design (a) is the least, followed by (b), (c) and (d). This aligns with the principle of $\wpd$, which is expected to have lowest value for null designs and highest for designs of the form $\Dfx$ (d). Moreover, note the relative differences in $\wpd$ values between (b) and (c). The value of the tuning parameter $\lambda$ is set to $0.67$, which has resulted in giving more emphasis to differences in x-axis categories. Again consider \ref{fig:onegran-new}(a) and \ref{fig:onegran-new}(b) with a $\wpd$ value of 20.5 and 145 respectively. This is because there is more gradual increase across hours of the day than months of the year. If order of the categories is not considered, they result in a $\wpd$ value of $97.8$ and $161$ respectively, which follows from the fact that if we consider difference between any hours of the day, the magnitude will be much larger than if we consider difference between consecutive categories. Similarly, \ref{fig:onegran-new}(a) and (b) have $\wpd$ values of  $110.79$ and $125.82$ respectively. The ranking implies that the distributional differences are more prominent for the second household, as is also seen from the bigger fluctuations in the $90^{th}$ percentile than the first household.

@@

@@

```{r raw, fig.cap = "(ref:raw)", fig.pos="!b"}
```

@@

(ref:dist-new-same-scale-link) The distribution of $\wpdsub{perm}$ and $\wpdsub{glm-scaled}$ are overlaid to compare the location and scale across different $\nx$ and $\nf$. $\wpdsub{norm}$ takes the value of $\wpdsub{perm}$ for lower levels, and $\wpdsub{glm-scaled}$ for higher levels to to alleviate the problem of computational time in permutation approaches. This is possible as the distribution of the adjusted measure looks similar for both approaches for higher levels.

```{r dist-new-same-scale-link, fig.cap = "(ref:dist-new-same-scale-link)"}
```
